State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms

### 2024-01-24 manuscript/manuscript_2/paper.tex
Final manuscript version

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source)

	 which also arise in traditional matched-filter analyses that exclude the pulsar terms.

 Including the pulsar terms also decreases the root-mean-square relative error in the estimated one-dimensional posteriors \textcolor{red}{by up to 99\%}, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quieter source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
\end{abstract}

@@

\begin{table*}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lccll}
			\toprule
			Set&Parameter & Injected value & Units & Prior  \\
			\hline
			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
			& $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
			& $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
			& $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
			& $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
			& $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
			& $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
			\hline
			\vspace{1mm}& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
      \multirow{2}{2mm}{$\boldsymbol{\theta'}_{\rm psr}$} & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right]  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
			\bottomrule
		\end{tabular}
		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar terms in Equation \eqref{eq:g_func_trig_chi}. The prior used for Bayesian inference is also displayed (rightmost column).  The top and bottom sections of the table contain $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta'}_{\rm psr}$ respectively. The subscript ``ATNF'' denotes values obtained from the ATNF pulsar catalogue as described in Section \ref{sec:rep_example1}. The subscript ``SC'' on $\sigma^{(n)}$ indicates that the injected value is calculated from Equation \eqref{eq:sigmap_f} and the empirical timing noise model for MSPs in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the uncertainties in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. We do not infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ for simplicity, so no prior is set. The priors on $\boldsymbol{\theta'}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
		}
		\label{tab:parameters_and_priors}
	\end{table*}
In order to synthesize $\boldsymbol{Y} = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$, which we convert to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. The numerical solutions depend on how we choose $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ or, equivalently, how we specify the configuration of a synthetic PTA. In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}$, namely $\boldsymbol{\theta}_{\rm gw}$. This latter step is equivalent to specifying the synthetic SMBHB source and differs from one test to the next according to the goal of the test.  \newline

@@

\begin{table}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lcll}
			\toprule
			&$\theta$ & $\Delta_{\rm Earth}(\theta)$ & $\Delta_{\rm psr \& earth}(\theta)$  \\
			\hline
			\multirow{7}{2mm}{High SNR} & $\Omega$       & $1.1 \times 10^{-2}$& $4.3 \times 10^{-3}$\\
			& $\Phi_0$ &$3.3 \times 10^{-3}$ &$1.2 \times 10^{-3}$ \\
			& $\psi$ &$9.7 \times 10^{-2}$ &$6.5 \times 10^{-4}$  \\
			& $\iota$ & $3.4 \times 10^{-1}$ & $1.0 \times 10^{-1}$  \\
			& $\delta$ &$1.9 \times10^{-3}$ &$2.9 \times 10^{-4}$  \\
			&$\alpha$ &$4.0 \times 10^{-2}$ &$4.4 \times10^{-4}$  \\
			&$h_0$ & $2.2 \times 10^{-1}$ &$7.5 \times 10^{-2}$  \\
			\hline
			\multirow{7}{2mm}{Low SNR} & $\Omega$       &  $2/7 \times 10^0$ &$1.6 \times 10^0$ \\
			& $\Phi_0$ &$3.4 \times 10^{-1}$ & $2.3 \times 10^{-1}$  \\
			& $\psi$ &$1.1 \times 10^{-1}$ &$1.2 \times 10^{-1}$  \\
			& $\iota$ & $4.4 \times 10^{-1}$ & $3.7 \times 10^{-1}$  \\
			& $\delta$ & $1.8 \times 10^{-1}$ & $7.8 \times 10^{-2}$  \\
			&$\alpha$ & $1.5 \times 10^{-1}$ & $1.4 \times 10^{-1}$ \\
			&$h_0$ & $1.1 \times 10^0$ &$1.4 \times 10^0$ \\
			\bottomrule
		\end{tabular}
		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=$\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$. The injected values are as summarised in Table \ref{tab:parameters_and_priors}. The top and bottom halves of the table contain the high-SNR ($h_0 =1 \times 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively. The psr\&Earth model is more accurate than the Earth model for all $\theta \in \boldsymbol{\theta}_{\rm gw}$ at high SNR and five out of seven parameters at low SNR.}
		\label{tab:posterior_errors}
	\end{table}

@@

The updated state-space model including the pulsar terms is tested on synthetic data. We consider 10 noise realisations for a single, astrophysically representative, SMBHB GW source with $h_0 = 5 \times 10^{-15}$, observed by the 12.5-year NANOGrav pulsars ($N=47$) with $T_{\rm obs} = 10 \, {\rm years}$ and $T_{\rm cad} = 1$ week. We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost. The irreducible random dispersion (cosmic variance) in the median of the marginalised one-dimensional posteriors, with a maximum $CV$ of 52\% for $\Phi_0$, a minimum of 0.2 \% for $\Omega$ and a median of $10 \%$ for $h_0$. The updated model is further tested across a broad and astrophysically plausible parameter domain of SMBHB sources, exploring 200 randomly sampled $\boldsymbol{\theta}_{\rm gw}$ at fixed $\iota$ and $h_0$. Consistent parameter estimates are obtained, although the accuracy is lower, when one or more PTA pulsars coincide approximately on the sky with the SMBHB source.For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases. For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases. Including the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) by 61\% for $\Omega$, up to 99\% for $\psi$ and $\alpha$ for a representative SMBHB system with a high SNR ($h_0 = 10^{-12}$). For low SNR ($h_0 = 5 \times 10^{-15}$), including the pulsar terms lowers the minimum detectable strain by $14\%$, comparable with standard PTA analyses \citep[e.g.][]]{Zhupulsarterms} \newline

@@

\section{Kalman filter} \label{sec:kalman}
The Kalman filter \citep{Kalman1} is a Gauss-Markov model used to algorithmically recover a temporal sequence of stochastically evolving  system state variables, $\boldsymbol{X}(t)$, which are not observed directly, given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$. It finds common use in engineering applications and has been applied successfully in neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}. In this appendix, we describe the Kalman filter algorithm used in this paper. General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the likelihood. The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.

\subsection{Recursion equations and likelihoood }\label{sec_kalman_general}
In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement). The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{X}(t_k)$. Each discrete timestep is indexed by $ 1 \leq k  \leq K$. The measurements are related to the states via
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter evolves the underlying states according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix. $\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process 
noise, with $\boldsymbol{w}_k \sim {\cal N}(0, \boldsymbol{Q}_k)$ and covariance $\boldsymbol{Q}_k$. \newline

@@

\section{Workflow summary}\label{sec:workflow}
For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
\begin{enumerate}[leftmargin=2em]
	\item Specify a PTA composed of $N$ pulsars 
	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$
	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
	\item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
	\item For each live point:
	\begin{enumerate}[leftmargin=2em]
		\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
	\end{enumerate}
	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
\end{enumerate}
In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$. The resulting $\mathcal{Z}$ values can then be compared. We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs. The generalization to TOAs is subtle and will be tackled in a forthcoming paper.

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha}
	\caption{Logarithm of the cross-section along the $\alpha$-axis of the likelihood function $\mathcal{L}_{\rm slice}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$. All static elements of $\boldsymbol{\theta}$ other than $\alpha$ are held fixed at their true injected values for the purpose of the tests in Appendix \ref{sec:psr_term_challenges}. In the left panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}_{\rm slice}(\alpha)$, i.e. inclusive of the pulsar terms. In the right-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig_earth}, i.e. just the Earth terms. The vertical dashed lines indicate the true injected value $\alpha = 1.0$ rad used to generate $\boldsymbol{Y}$. The inset in the left panel covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad. In the left panel $\mathcal{L}_{\rm slice}(\alpha)$ is jagged, although a unique peak exists in principle at $\alpha=1.0 \, {\rm rad}$, and $\mathcal{L}_{\rm slice}(\alpha)$ is smooth within the neighbourhood plotted in the inset. The jagged form of $\mathcal{L}_{\rm slice}(\alpha)$ hampers the convergence of the nested sampler. In the right panel $\mathcal{L}_{\rm slice}(\alpha)$ is smooth for $0\leq\alpha\leq \pi$. }
	\label{fig:likelihood_comparison}
\end{figure*}
In this appendix we elucidate the complications introduced by the pulsar terms into the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}. \newline

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) which also arise in traditional matched-filter analyses that exclude the pulsar terms. Additionally, including the pulsar terms decreases the minimum detectable strain by $14\%$.
\end{abstract}

@@

\begin{table*}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lccll}
			\toprule
			Set&Parameter & Injected value & Units & Prior  \\
			\hline
			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
			& $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
			& $\delta$              & $1.0$  & rad & Cosine($-\pi/2, \pi/2$) \\
			& $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
			& $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
			& $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
			& $\iota$             & $1.0$ & rad & Sin($0, \pi$) \\ 
			\hline
			\vspace{1mm}& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
      \multirow{2}{2mm}{$\boldsymbol{\theta'}_{\rm psr}$} & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right]  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
			\bottomrule
		\end{tabular}
		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar terms in Equation \eqref{eq:g_func_trig_chi}. The prior used for Bayesian inference is also displayed (rightmost column).  The top and bottom sections of the table contain $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta'}_{\rm psr}$ respectively. The subscript ``ATNF'' denotes values obtained from the ATNF pulsar catalogue as described in Section \ref{sec:rep_example1}. The subscript ``SC'' on $\sigma^{(n)}$ indicates that the injected value is calculated from Equation \eqref{eq:sigmap_f} and the empirical timing noise model for MSPs in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the uncertainties in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. We do not infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ for simplicity, so no prior is set. The priors on $\boldsymbol{\theta'}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
		}
		\label{tab:parameters_and_priors}
	\end{table*}
In order to synthesize $\boldsymbol{Y} = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$, which we convert to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. The numerical solutions depend on how we choose $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ or, equivalently, how we specify the configuration of a synthetic PTA. In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}$, namely $\boldsymbol{\theta}_{\rm gw}$. This latter step is equivalent to specifying the synthetic SMBHB source and differs from one test to the next according to the goal of the test.  \newline

@@

\begin{table}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lcll}
			\toprule
			&$\theta$ & $\Delta_{\rm Earth}(\theta)$ & $\Delta_{\rm psr \& Earth}(\theta)$  \\
			\hline
			\multirow{7}{2mm}{High SNR} & $\Omega$       & $1.1 \times 10^{-2}$& $4.3 \times 10^{-3}$\\
			& $\Phi_0$ &$3.3 \times 10^{-3}$ &$1.2 \times 10^{-3}$ \\
			& $\psi$ &$9.7 \times 10^{-2}$ &$6.5 \times 10^{-4}$  \\
			& $\iota$ & $3.4 \times 10^{-1}$ & $1.0 \times 10^{-1}$  \\
			& $\delta$ &$1.9 \times10^{-3}$ &$2.9 \times 10^{-4}$  \\
			&$\alpha$ &$4.0 \times 10^{-2}$ &$4.4 \times10^{-4}$  \\
			&$h_0$ & $2.2 \times 10^{-1}$ &$7.5 \times 10^{-2}$  \\
			\hline
			\multirow{7}{2mm}{Low SNR} & $\Omega$       &  $2/7 \times 10^0$ &$1.6 \times 10^0$ \\
			& $\Phi_0$ &$3.4 \times 10^{-1}$ & $2.3 \times 10^{-1}$  \\
			& $\psi$ &$1.1 \times 10^{-1}$ &$1.2 \times 10^{-1}$  \\
			& $\iota$ & $4.4 \times 10^{-1}$ & $3.7 \times 10^{-1}$  \\
			& $\delta$ & $1.8 \times 10^{-1}$ & $7.8 \times 10^{-2}$  \\
			&$\alpha$ & $1.5 \times 10^{-1}$ & $1.4 \times 10^{-1}$ \\
			&$h_0$ & $1.1 \times 10^0$ &$1.4 \times 10^0$ \\
			\bottomrule
		\end{tabular}
		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = Earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=$\rm psr \& Earth$, Equation \eqref{eq:g_func_trig}) for $\theta \in \boldsymbol{\theta}_{\rm gw}$. The injected values are as summarised in Table \ref{tab:parameters_and_priors}. The top and bottom halves of the table contain the high-SNR ($h_0 =1 \times 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively. The psr\&Earth model is more accurate than the Earth model for all $\theta \in \boldsymbol{\theta}_{\rm gw}$ at high SNR and five out of seven parameters at low SNR.}
		\label{tab:posterior_errors}
	\end{table}

@@

The updated state-space model including the pulsar terms is tested on synthetic data. We consider 10 noise realisations for a single, astrophysically representative, SMBHB GW source with $h_0 = 5 \times 10^{-15}$, observed by the 12.5-year NANOGrav pulsars ($N=47$) with $T_{\rm obs} = 10 \, {\rm years}$ and $T_{\rm cad} = 1$ week. We find that the updated state-space model successfully detects injected signals and estimates their static parameters accurately for all noise realisations with relatively low computational cost. The irreducible random dispersion (cosmic variance) in the median of the marginalised one-dimensional posteriors, with a maximum $CV$ of 52\% for $\Phi_0$, a minimum of 0.2 \% for $\Omega$ and a median of $10 \%$ for $h_0$. The updated model is further tested across a broad and astrophysically plausible parameter domain of SMBHB sources, exploring 200 randomly sampled $\boldsymbol{\theta}_{\rm gw}$ at fixed $\iota$ and $h_0$. Consistent parameter estimates are obtained, although the accuracy is lower, when one or more PTA pulsars coincide approximately on the sky with the SMBHB source.For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases. For the worst estimated parameter $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases. Including the pulsar terms increases the accuracy (as quantified by the root-mean-square relative error, Equation \eqref{eq:mean_rel_error}) for five out of the seven static parameters in $\boldsymbol{\theta}_{\rm gw}$; we find $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for all parameters except $\psi$ and $h_0$. In the high-SNR case ($h_0 = 1 \times 10^{-12}$) all of the seven static parameters are estimated more accurately; $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$. Including the pulsar terms lowers the minimum detectable strain by $14\%$, comparable with standard PTA analyses \citep[e.g.][]{Zhupulsarterms}. \newline

@@

\section{Kalman filter} \label{sec:kalman}
The Kalman filter \citep{Kalman1} is an algorithmic optimal solver for Gauss-Markov processes. Given a temporal sequence of noisy measurements, $\boldsymbol{Y}(t)$ the Kalman filter recovers a temporal sequence of stochastically evolving system state variables, $\boldsymbol{X}(t)$, which are hidden from the observer. It is a standard method in control theory \citep{zarchan2000fundamentals,rob_KF_book,10.5555/3103280} and finds common use in engineering applications \citep[e.g.][]{KFexample1,KFexample2,KFexample3,KFexample4} as well as successful application to neutron star astrophysics \citep[e.g.][]{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}. In this appendix, we outline the Kalman filter used in this paper. The general recursion relations for the discrete-time Kalman filter are presented for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}, along with the formula for the likelihood. The application of the Kalman filter algorithm to the specific continuous-time state-space model in Section \ref{sec:2} is outlined in Section \ref{sec_kalman_specific}.

\subsection{Recursion equations and likelihood }\label{sec_kalman_general}
In this work we use the linear Kalman filter, which assumes a linear relation between $d{\boldsymbol{X}}/dt$ and ${\boldsymbol{X}}(t)$ (dynamics) and between ${\boldsymbol{Y}}(t)$ and ${\boldsymbol{X}}(t)$ (measurement). The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k = \boldsymbol{Y}(t_k)$, which are related via a linear transformation to a set of unobservable discrete system states $\boldsymbol{X}_k= \boldsymbol{X}(t_k)$. Each discrete timestep is indexed by $ 1 \leq k  \leq K$. The measurements are related to the states via
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter evolves the underlying states according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix. $\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process 
noise, with $\boldsymbol{w}_k \sim {\cal N}(0, \boldsymbol{Q}_k)$ and covariance $\boldsymbol{Q}_k$. \newline

@@

\section{Workflow summary}\label{sec:workflow}
For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
\begin{enumerate}[leftmargin=2em]
	\item Specify a PTA composed of $N$ pulsars 
	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$
	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
	\item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
	\item For each live point:
	\begin{enumerate}[leftmargin=2em]
		\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
		\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value through Equation \eqref{eq:likelihood}
	\end{enumerate}
	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
\end{enumerate}
In order to compute the odds ratio $\beta$ the above workflow is repeated for a different $\mathcal{M}$. The resulting $\mathcal{Z}$ values can then be compared. We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs. The generalization to TOAs is subtle and will be tackled in a forthcoming paper.

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth ]{images/Likelihoods_alpha_v3}
	\caption{Logarithm of the cross-section along the $\alpha$-axis of the likelihood function $\mathcal{L}_{\rm slice}(\alpha)$ returned by the Kalman filter acting on synthetic data $\boldsymbol{Y}$. All static elements of $\boldsymbol{\theta}$ other than $\alpha$ are held fixed at their true injected values for the purpose of the tests in Appendix \ref{sec:psr_term_challenges}. In the left panel the Kalman filter uses Equation \eqref{eq:g_func_trig} to calculate $\mathcal{L}_{\rm slice}(\alpha)$, i.e. inclusive of the pulsar terms. In the right-hand panel the Kalman filter uses Equation \eqref{eq:g_func_trig_earth}, i.e. just the Earth terms. The vertical dashed lines indicate the true injected value $\alpha = 1.0$ rad used to generate $\boldsymbol{Y}$. The inset in the left panel covers the region $1.0 - 10^{-4} \le \alpha \le 1.0 +10^{-4}$ rad. In the left panel $\mathcal{L}_{\rm slice}(\alpha)$ is jagged, although a unique peak exists in principle at $\alpha=1.0 \, {\rm rad}$, and $\mathcal{L}_{\rm slice}(\alpha)$ is smooth within the neighbourhood plotted in the inset. The jagged form of $\mathcal{L}_{\rm slice}(\alpha)$ hampers the convergence of the nested sampler. In the right panel $\mathcal{L}_{\rm slice}(\alpha)$ is smooth for $0\leq\alpha\leq \pi$. The green (blue) curve is the likelihood returned by $\mathcal{M}_{\rm psr\&Earth}$ ($\mathcal{M}_{\rm Earth}$).}
	\label{fig:likelihood_comparison}
\end{figure*}
In this appendix we elucidate the complications introduced by the pulsar terms into the state-space analysis and justify the reparametrisation in terms of $\chi^{(n)}$ introduced in Section \ref{sec:pulsar_term2}. \newline
